---
title: "Analysis of severe weather events impacting population health and the economy in the U.S."
author: "Ralph Waldenmaier"
date: "Sunday, March 22, 2015"
output: html_document
---

# Synopsis

The U.S. National Oceanic and Atmospheric Administration (NOAA) collects data about storms and weather-related events from around the country on a near-daily basis. This includes the number of fatalities and injuries, along with amount of property damage and crop damage. 
This report analyzes the data to determine the types of events that are most harmful with respect to population health, and which types of events have the greatest economic consequences. The analisys shows that tornadoes cause the largest number of injuries and death, while floods are responsible for the largest total economic cost.

# Data Processing

Here is brief overview about the data processing in this analysis

1. Downloading the NOAA data using a link provided with the Coursera assessment.
2. Loading the data to memory using the read.csv and bzip2 functions
3. Cleaning the data set to remove unnecessary duplicates for instance due to typos
4. Cleaning the data sets calculation values
5. Preparing result sets for the graph generation
6. Graph generation to address the required questions.

```{r}
knitr::opts_chunk$set(cache=TRUE, echo = TRUE, warning = FALSE, message = FALSE)
```

First of all the required libraries for this analysis are loaded.
```{r Load libraries, cache=TRUE }
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(ggplot2)
library(scales)
library(lubridate)
```

The data provided by Coursera is obtained and stored on disk.
```{r Obtain NOAA sotrm data, cache=TRUE}
stormUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if(!file.exists("repdata_data_StormData.csv.bz2")) 
{
  download.file(stormUrl, destfile = "repdata_data_StormData.csv.bz2")  
}
```

In order to save some disk space, the data is directly extracted from the bzip2 file and loaded to memory using the read.csv function.

```{r Data Processing, cache=TRUE }
zipfile <- "repdata_data_StormData.csv.bz2"
if(!file.exists(zipfile)) 
{
  print("error: file does not exist")
}
data <- tbl_df(read.csv(bzfile(zipfile),
                          stringsAsFactors=FALSE, 
                          header=TRUE
                        )
               )
rm(zipfile)
```

## Preliminary Data Cleaning

The data is cleaned in several steps.
First of all, the values that do not provide any values(zeroes) are filtered out from the original data set. Additionally for the events cleanup, the EVTYPE is unified using tolower producing lowercase letters.
```{r Cleaning 1 ,cache=TRUE}
cleanData <- data %>% 
              filter(FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0) %>%
              mutate(EVTYPE=tolower(EVTYPE))
```

Second, the EVTYPE needs to be unified. Here, several processing s based on the document [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) are executed.
```{r Cleaning 2 ,cache=TRUE}
cleanData <- cleanData %>%
              mutate(EVTYPE=gsub("[[:punct:]]", "", EVTYPE)) %>%
              mutate(EVTYPE=gsub("^\\s+", "", EVTYPE)) %>% # removing whitespace in the beginning
              mutate(EVTYPE=gsub("\\s+&", "", EVTYPE)) %>% # removing whitespace in the end
              filter(nchar(EVTYPE) > 0) %>% # removing empty events
              mutate(EVTYPE=gsub("astron", "AstronomicalLowTide", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("aval", "Avalanche", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("bliz", "Blizzard", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("(beach|coast).*flood", "CoastalFlood", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*extreme.*)(?=.*(cold|wind).*chill).*$", "ColdWindChill", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("debris", "DebrisFlow", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^dense fog", "DenseFog", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^dense smoke", "DenseSmoke", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("drought", "Drought", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*waterspout.*)(?=.*dust dev).*$", "DustDevil", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("dust.*storm", "DustStorm", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("excess.*heat", "ExcessiveHeat", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("extreme.*(cold|wind)", "ExtremeColdWindChill", EVTYPE, perl=TRUE)) %>%
              #mutate(EVTYPE=gsub("flash.*flood.*", "FlashFlood", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*flash.*)(?=.*flood.*|fld.*).*$", "Flood", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("frost|freeze", "FrostFreeze", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*waterspout.*)(?=.*funnel).*$", "FunnelCloud", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("freez.*fog", "FreezingFog", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*marine.*)(?=.*hail).*$", "Hail", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*excessive.*)(?=.*heat).*$", "Heat", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("heavy.*rain", "HeavyRain", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("heavy.*snow", "HeavySnow", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("high.*surf", "HighSurf", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*marine.*)(?=.*high.*wind).*$", "HighWind", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("hurric|typhoon", "HurricaneTyphoon", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("ice.*storm", "IceStorm", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("lake.*effect.*snow", "LakeEffectSnow", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("shore.*flood", "LakeshoreFlood", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("lightning|lighting", "Lightning", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("marine hail", "MarineHail", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("marine high wind", "MarineHighWind", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("marine strong wind", "MarineStrongWind", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("marine.*(thunder|tstm)", "MarineThunderstormWind", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("rip.*current", "RipCurrent", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("seiche", "Seiche", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("sleet", "Sleet", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("storm.*(surge|tide)", "StormSurgeTide", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*marine.*)(?=.*strong.*wind).*$", "StrongWind", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*marine|non tstm|non-tstm.*)(?=.*(tstm|thunde|tunder).*(wind|wins)).*$", "ThunderstormWind", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("^(?!.*waterspout.*)(?=.*torn).*$", "Tornado", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("trop.*depr", "TropicalDepression", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("trop.*storm", "TropicalStorm", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("tsunami", "Tsunami", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("volc", "VolcanicAsh", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("(water|wayter).*spout", "Waterspout", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("wild.*fire", "Wildfire", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("winter.*storm", "WinterStorm", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("winter.*weather", "WinterWeather", EVTYPE, perl=TRUE)) %>%
              mutate(EVTYPE=gsub("HurricaneTyphoon.*", "HurricaneTyphoon", EVTYPE, perl=TRUE))            
              
```

Besides the EVTYPE (event type), the calculation values (PROPDMG[Property damage] and CROPDMG[Crop damage]) need to be cleaned. Within the variables (PROPDMGEXP and CROPDMGEXP) are the respective multipliers encoded which need to be translated and applied to (PROPDMG and CROPDMG) variables.
```{r Cleaning 3, cache=TRUE}
#x <- cleanData %>% select(EVTYPE) %>% distinct() # remove this after debugging
#x

# Cleaning property damage data
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="0", PROPDMG * 1, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="1", PROPDMG * 10, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="2", PROPDMG * 100, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="3", PROPDMG * 1000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="4", PROPDMG * 10000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="5", PROPDMG * 100000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="6", PROPDMG * 1000000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="7", PROPDMG * 10000000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="8", PROPDMG * 100000000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="B", PROPDMG * 1000000000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="b", PROPDMG * 1000000000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="H", PROPDMG * 100, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="h", PROPDMG * 100, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="K", PROPDMG * 1000, PROPDMG))
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="k", PROPDMG * 1000, PROPDMG))  
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="M", PROPDMG * 1000000, PROPDMG))  
cleanData <- cleanData %>% mutate(PROPDMG=ifelse(PROPDMGEXP=="m", PROPDMG * 1000000, PROPDMG))  

# Cleaning crop damage data
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="0", CROPDMG * 1, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="1", CROPDMG * 10, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="2", CROPDMG * 100, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="3", CROPDMG * 1000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="4", CROPDMG * 10000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="5", CROPDMG * 100000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="6", CROPDMG * 1000000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="7", CROPDMG * 10000000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="8", CROPDMG * 100000000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="B", CROPDMG * 1000000000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="b", CROPDMG * 1000000000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="H", CROPDMG * 100, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="h", CROPDMG * 100, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="K", CROPDMG * 1000, CROPDMG))
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="k", CROPDMG * 1000, CROPDMG))  
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="M", CROPDMG * 1000000, CROPDMG))  
cleanData <- cleanData %>% mutate(CROPDMG=ifelse(CROPDMGEXP=="m", CROPDMG * 1000000, CROPDMG))  

```

# Results

## Result data processing
Using the previously cleaned data, the result sets for the needed graphs can be generated.

First the data for the causalities is calculated. Therefore the data is grouped by EVTYPE and the values for INJURIES and FATALITIES are summed. After adding INJURIES and FATALITIES, a new variable impact is generated per EVTYPE. To provide only the most interesting values to the result, only the 15 records with the highest impact are chosen.

```{r Resultset generation causalities, cache=TRUE}
causalities <- cleanData %>% 
                group_by(EVTYPE) %>% 
                summarize(inj=sum(INJURIES), fat=sum(FATALITIES)) %>% 
                mutate(impact=inj+fat) %>% 
                arrange(desc(impact)) %>%
                top_n(15, impact)
```


The same calculation as previously must be generated for the economic impact as well. For this, the values for COPRDMG and PROPDMG, which were cleaned in the previous steps, are now summarized after grouping them by EVTYPE. Also here an impact is generated, though divided by one million in order to have more readable values in the charts using the grid. Also here only the 15 records with the highest impact are chosen.

```{r Resultset generation economic, cache=TRUE}
economic <- cleanData %>% 
                group_by(EVTYPE) %>% 
                summarize(crp=sum(CROPDMG), prp=sum(PROPDMG)) %>% 
                mutate(impact=(crp+prp)/1000000) %>% 
                arrange(desc(impact)) %>%
                top_n(15, impact)
```

## Graphs

## Answering question 1

The most harmful event with respect to population health are the **tornadoes**

```{r Generating graph causalities, cache=TRUE}
ggplot(causalities, aes(x=reorder(EVTYPE,impact), y=impact)) +
        geom_bar(stat="identity") +
        xlab("Event type") +
        ylab("# of victims") +
        coord_flip() +
        ggtitle("Most harmful events with respect to population health")
```

## Answering question 2

The event with the greatest economic consequences is the **flood**

```{r Generating graph economic, cache=TRUE}
ggplot(economic, aes(x=reorder(EVTYPE,impact), y=impact)) +
        geom_bar(stat="identity") +
        xlab("Event type") +
        ylab("Costs in millions") +
        coord_flip() +
        ggtitle("Economic consequences by event type") +
        scale_y_continuous(labels = comma) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

